---
title: "BDA_project"
output: pdf_document
Authors: "Edvard Ohlström", "Adam Harmat", "Yeralkhan Slam"
date: "2023-11-06"
---

```{r setup, include=FALSE}
if(!require(medicaldata)){
  install.packages("medicaldata")
}

library(medicaldata)
library(aaltobda)
library(bayesplot)
library(cmdstanr)
library(dplyr)
library(ggplot2)
library(ggdist)
library(posterior)

```

## R Markdown

Dataset description -  
https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/codebooks/indo_rct_codebook.html



columns 1
Outcome = Wheter they have the dises X

```{r}
data(strep_tb)
current_dir<-getwd()
filename<-paste(current_dir,"/strep_tb.csv")
filename<-gsub(" ", "", filename)
write.csv(strep_tb,filename)

#Put numeric values to binomial data. True=1, False=0
df_numeric <- sterp_tb%>%
mutate(across(where(is.factor), ~ as.numeric(.x == "True")))
```





Step 1. 
 - EDA , goal: Explore
Step 2
 - based on earlier results choose methods.

Notes:

R1. How good is the drug/treatment for baseline condition/?
R2. When is the drug worth giving to patients?

FAR out
R3. What is the alternative cost for society and 


Assume all hospitals have same distribution vs individual (non hierarchical and hierarchical)
Test risk group 1, risk group 2...  (variable selection with many models)