---
title: "BDA_project"
output: pdf_document
Authors: "Edvard Ohlström", "Adam Harmat", "Yeralkhan Slam"
date: "2023-11-06"
---

```{r setup, include=FALSE}
if(!require(medicaldata)){
  install.packages("medicaldata")
}

library(medicaldata)
library(aaltobda)
library(bayesplot)
library(cmdstanr)
library(dplyr)
library(ggplot2)
library(ggdist)
library(posterior)
library(dplyr)
library(reshape2)
library(rstan)
library(cmdstanr)
library(aaltobda)
library(bayesplot)
library(cmdstanr)
library(dplyr)
library(ggdist) # for stat_dotsinterval
library(posterior)
```

## R Markdown

Dataset description -  https://htmlpreview.github.io/?https://github.com/higgi13425/medicaldata/blob/master/man/description_docs/strep_tb_desc.html 


```{r}

data("strep_tb")
strep_tb <- na.omit(strep_tb)
summary(strep_tb)
data <- read.csv(file = "strep_tb_scaled.csv", header = TRUE)
data$X <- NULL
data$n = 1

```


Investigate the drug's efficacy based on different patient baselines, particularly focusing on whether it is more effective for individuals with severe illness compared to those less sick.

you have x y z... features => will the drug work?

find function that gives the prediction based on the patients base conditions

1. logistic regression | input(x|theta, y|theta, z|theta) ->  probablility  {0,1} per patient | model assumption the features are not correlated

  Case 1 - no separation (joint model)
  Case 2 - group by baseline condition (seperate model)
  Case 3 - Do not group but find the common (hierarichal model)

logistic hierarichal | input(x|theta1, y|theta2, z|theta3 ) -> {0,1} per patient | model assumption the feature correlated




```{r}

matrix <- cbind(data$baseline_temp, data$baseline_esr)
y_pred = 1:length(data$baseline_temp)

data_list <- list(
  N = length(data$baseline_temp),
  X = matrix,
  y = data$improved
  )



#fit <- stan(file = "logistic_model.stan", data = data_list, chains = 4, iter = 2000)

```


TEMP Beta1 prior mean = (y2-y-1)  (1-0)/(101-99) = 0.5
TEMP Beta1 prior SD = (diff(extreme value,TEMP prior mean )  - TEMP prior mean ) / 3 = (1-0.5) /3 = 0,1666666667 # probably not good

ESR prior mean = 52-15.5
ESR prior sd = 
````{r}
 m = 1/(52-15.5)
sd = 1/()
````


````{r}
model_pooled<- cmdstan_model(stan_file = "logistic_model.stan")

# Sampling from the posterior distribution happens here:
fit<- model_pooled$sample(data = data_list, refresh=0,
                                      show_messages=FALSE,
                                      show_exceptions=FALSE)
print(fit)

```

```{r}
#generated_values <- extract(fit)

#fit$cmdstan_diagnose()
mean_values <- fit$summary()

prob_mean <- mean_values$mean[5:length(mean_values$mean)]

output <- cbind(prob_mean, data$improved)

prob_mean_drug <- prob_mean[1:53]
probl_mean_placebo <-  prob_mean[54:length(prob_mean)]

hist(prob_mean_drug, )
hist(probl_mean_placebo)
unique(prob_mean)
```

```{r}

cbind(y_pred_prob, data)

```

Assume all hospitals have same distribution vs individual (non hierarchical and hierarchical)
Test risk group 1, risk group 2...  (variable selection with many models)
